---
title: "PCA: Group 03"
format:
  html:
    embed-resources: true
editor: visual
---

### Import dataset

```{r,message=FALSE,warning=FALSE}
library(tidyverse)
library(broom)
library(cowplot)
library(ggplot2)
library(here)
aug_data <- read_csv(here("data/03_dat_aug.csv"))
```

### Start of the PCA

Looking at the data in PC coordinates:

```{r pc_coordinate_plot}
numeric_data <- aug_data |> 
  select(where(is.numeric)) |> 
  na.omit() |> 
  select(where(~ var(., na.rm = TRUE) > 0))

pca_fit <- numeric_data |>  
  prcomp(scale = TRUE) 

data_clean <- aug_data |> 
  filter(row_number() %in% rownames(numeric_data))

pca_plot <- pca_fit |> 
  augment(data_clean) |>  
  ggplot(aes(.fittedPC1, .fittedPC2)) + 
  geom_point(size = 1.5) +
  theme_half_open(12) +
  background_grid()
pca_plot
```

Now we extract the rotation matrix

```{r}
pca_fit |> 
  tidy(matrix = "rotation")
```

and plot it:

```{r rotation_matrix_plot}
# arrow for plotting
arrow_style <- arrow(
  angle = 20, ends = "first", type = "closed", length = grid::unit(8, "pt")
)

# rotation matrix
pca_fit |> 
  tidy(matrix = "rotation") |> 
  pivot_wider(names_from = "PC",
              names_prefix = "PC",
              values_from = "value") |> 
  ggplot(aes(PC1, PC2)) +
  geom_segment(xend = 0,
               yend = 0,
               arrow = arrow_style) +
  geom_text(
    aes(label = column),
    hjust = 1, nudge_x = -0.02, 
    color = "#904C2F"
  ) +
  xlim(-1.25, .5) +
  ylim(-.5, 1) +
  coord_fixed() + 
  theme_minimal_hgrid(12)
```

Look at the variance

```{r}
pca_fit |> 
  tidy(matrix = "eigenvalues")
```

And now the plot:

```{r variance_plot}
pca_bar_plot <- pca_fit |> 
  tidy(matrix = "eigenvalues") |> 
  ggplot(aes(PC, percent)) +
  geom_col(fill = "#56B4E9",
           alpha = 0.8) +
  scale_x_continuous(breaks = 1:9) +
  scale_y_continuous(
    labels = scales::percent_format(),
    expand = expansion(mult = c(0, 0.01))
  ) +
  theme_minimal_hgrid(12)
pca_bar_plot
```

It is seen that 22.5% of the variance is explained by PC1. The drop of is after PC3.

### Exporting the key plots

```{r}
ggsave(
  filename = here("results", "06_pca_plot.png"),
  plot = pca_plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)

ggsave(
  filename = here("results", "06_pca_bar_plot.png"),  
  plot = pca_bar_plot, 
  width = 8, 
  height = 6, 
  dpi = 300
)
```

We can then extract what each Principle component is made up of:

```{r}
# Extract the loadings
loadings <- pca_fit$rotation |> 
  as_tibble(rownames = "Variable")

print(loadings)
```

#### Using k-means to cluster patients

```{r}
pca_data <- pca_fit |> 
  augment(data_clean)

pca_data <- pca_data |> 
  mutate(cluster = kmeans(select(pca_data, .fittedPC1:.fittedPC3), centers = 3)$cluster) |>
  mutate(cluster = as.factor(cluster))

pca_clustered <- pca_data |> 
  ggplot(aes(x = .fittedPC1, y = .fittedPC2, color = cluster)) +
  labs(x = "PC1 - (encounters)",
       y = "PC2 - (time_in_hospital)",
       title = "Clustering of patient groups based on PCA") +
  geom_point() +
  theme_minimal()
```

### Exporting the key plot again

```{r}
ggsave(
  filename = here("results", "06_pca_clustered.jpg"),
  plot = pca_clustered , 
  width = 8, 
  height = 6, 
  dpi = 300
)
```

### Summary

-   **PC1**: Likely represents overall patient interaction with the healthcare system.

-   **PC2**: Likely represents treatment intensity or complexity.

-   **PC3**: Likely represents inpatient care or hospitalization frequency.
